---
title: "Chapter 14"
author: "Aditya Dahiya"
subtitle: "Numbers"
date: 2023-09-08
execute: 
  echo: true
  warning: false
  error: false
  cache: true
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/RfDS2solutions/Chapter14.html"
  description: "Solutions: R for Data Science (2e)"
  twitter: true
  facebook: true
  linkedin: true
  email: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r}
library(tidyverse)
library(gt)
library(nycflights13)
data("flights")
```

# **14.3.1 Exercises**

## Question 1

**How can you use [`count()`](https://dplyr.tidyverse.org/reference/count.html) to count the number rows with a missing value for a given variable?**

We can use `count()` with weights as `is.na(var_name)` to the number of rows with missing values, as shown below

```{r}
#| layout-ncol: 2

flights |>
  group_by(month) |>
  summarise(total = n(),
            missing = sum(is.na(dep_time))) |>
  gt()


flights |>
  group_by(month) |>
  count(wt = is.na(dep_time)) |>
  ungroup() |>
  gt()
```

## Question 2

Expand the following calls to [`count()`](https://dplyr.tidyverse.org/reference/count.html) to instead use [`group_by()`](https://dplyr.tidyverse.org/reference/group_by.html), [`summarize()`](https://dplyr.tidyverse.org/reference/summarise.html), and [`arrange()`](https://dplyr.tidyverse.org/reference/arrange.html):

1.  **`flights |> count(dest, sort = TRUE)`**

    ```{r}
    flights |>
      group_by(dest) |>
      summarise(n = n()) |>
      arrange(desc(n))
    ```

2.  **`flights |> count(tailnum, wt = distance)`**

    ```{r}
    flights |>
      group_by(tailnum) |>
      summarise(n = sum(distance))
    ```

# **14.4.8 Exercises**

## Question 1

**Explain in words what each line of the code used to generate [Figure 14.1](https://r4ds.hadley.nz/numbers#fig-prop-cancelled) does.**

The explanation is given as annotation in the code below (i.e, lines starting with `#`): --

```{r}
# Load in the data-set flights from package nycflights13
flights |> 
  # Create a variable hour, which is the quotient of the division of
  # sched_dep_time by 100. Further, group the dataset by this newly 
  # created variable "hour" to get data into 24 groups - one for each
  # hour.
  group_by(hour = sched_dep_time %/% 100) |> 
  
  # For each gropu, i.e. all flights scheduled to depart in that
  # hour, compute the NAs, i.e. cancelled flights, then compute their
  # mean, i.e. proportion of cancelled flights; and also create a 
  # variable n, which is the total number of flights
  summarize(prop_cancelled = mean(is.na(dep_time)), n = n()) |> 
  
  # Remove the flights departing between 12 midnight and 1 am, since
  # these are very very few, and all are cancelled leading to a highly
  # skewed and uninformative graph
  filter(hour > 1) |> 
  
  # Start a ggplot call, plotting the hour on the x-axis, and 
  # proportion of cancelled flights on the y-axis
  ggplot(aes(x = hour, y = prop_cancelled)) +
  
  # Create a line graph,which joins the proportion of cancelled 
  # flights for each hour. Also, put in in dark grey colour
  geom_line(color = "grey50") + 
  
  # Add points for each hour, whose size varies with the total number
  # of flights in that hour
  geom_point(aes(size = n))
```

## Question 2

**What trigonometric functions does R provide? Guess some names and look up the documentation. Do they use degrees or radians?**

`R` provides several trigonometric functions, most of which operate in radians. Here's a table listing some of the commonly used trigonometric functions in R, along with short descriptions and information about whether they use degrees or radians:

| **Function**               | **Description**                                                                                                                | **Angle Measure** |
|--------------|--------------------------------------------|--------------|
| `sin(x)`                   | Sine function: Computes the sine of the angle `x`.                                                                             | Radians           |
| `cos(x)`                   | Cosine function: Computes the cosine of the angle `x`.                                                                         | Radians           |
| `tan(x)`                   | Tangent function: Computes the tangent of the angle `x`.                                                                       | Radians           |
| `asin(x)` or `acos(x)`     | Inverse sine or inverse cosine: Computes the angle whose sine or cosine is `x`.                                                | Radians           |
| `atan(x)` or `atan2(y, x)` | Inverse tangent or arctangent: Computes the angle whose tangent is `x` or the angle between the point `(x, y)` and the origin. | Radians           |
| `sinh(x)`                  | Hyperbolic sine function: Computes the hyperbolic sine of `x`.                                                                 | Radians           |
| `cosh(x)`                  | Hyperbolic cosine function: Computes the hyperbolic cosine of `x`.                                                             | Radians           |
| `tanh(x)`                  | Hyperbolic tangent function: Computes the hyperbolic tangent of `x`.                                                           | Radians           |

In R, trigonometric functions like `sin`, `cos`, and `tan` expect angles to be in radians by default. However, we can convert between degrees and radians using the `deg2rad` and `rad2deg` functions. For example, to compute the sine of an angle in degrees, you can use `sin(deg2rad(angle))`.

```{r}
#| label: fig-q2-ex4
#| fig-cap: Graphs from some common trigonometric functions
#| fig-width: 10
df = tibble(x = seq(from = -5, to = +5, by = 0.1))
g = ggplot(df, aes(x = x)) + 
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = NULL, y = NULL) +
  scale_x_continuous(breaks = -5:5)
gridExtra::grid.arrange(
  g + geom_line(aes(y = sin(x))) + labs(title = "sin(x)"),
  g + geom_line(aes(y = cos(x))) + labs(title = "cos(x)"),
  g + geom_line(aes(y = tan(x))) + labs(title = "tan(x)"),
  g + geom_line(aes(y = asin(x))) + labs(title = "asin(x)"),
  g + geom_line(aes(y = acos(x))) + labs(title = "acos(x)"),
  g + geom_line(aes(y = atan(x))) + labs(title = "atan(x)"),
  g + geom_line(aes(y = sinh(x))) + labs(title = "sinh(x)"),
  g + geom_line(aes(y = cosh(x))) + labs(title = "cosh(x)"),
  
  nrow = 2
)
```

## Question 3

**Currently `dep_time` and `sched_dep_time` are convenient to look at, but hard to compute with because they\'re not really continuous numbers. You can see the basic problem by running the code below: there\'s a gap between each hour.**

```         
flights |>    
  filter(month == 1, day == 1) |>    
  ggplot(aes(x = sched_dep_time, y = dep_delay)) +   
  geom_point()
```

**Convert them to a more truthful representation of time (either fractional hours or minutes since midnight).**

We can correct the `sched_dep_time()` using the two arithmetical functions `%/%` and `%%` to generate the decimal representation of time in hours as shown in the code for the graph on the right hand side in

```{r}
#| label: fig-q3-ex4
#| fig-cap: "Improved represenation of scheduled departure time to remove breaks in the data owing to represention of time as hhmm"

gridExtra::grid.arrange(
  flights |>    
    filter(month == 1, day == 1) |>    
    ggplot(aes(x = sched_dep_time, y = dep_delay)) +   
    geom_point() +
    labs(subtitle = "Incorrect scheduled departure time"),

  flights |>
    mutate(
      hour_dep = sched_dep_time %/% 100,
      min_dep  = sched_dep_time %%  100,
      time_dep = hour_dep + (min_dep/60)
    ) |>
    filter(month == 1, day == 1) |>    
    ggplot(aes(x = time_dep, y = dep_delay)) +   
    geom_point() +
    labs(subtitle = "Improved and accurate scheduled departure time",
         x = "Scheduled Departure Time (in hrs)") +
    scale_x_continuous(breaks = seq(0,24,4)),
  
  ncol = 2)
```

## Question 4

**Round `dep_time` and `arr_time` to the nearest five minutes.**

```{r}
attach(flights)
flights |>
  slice_head(n = 50) |>
  mutate(
    dep_time_5 = round(dep_time/5) * 5,
    arr_time_5 = round(arr_time/5) * 5,
    .keep = "used"
  ) |>
  gt() |>
  opt_interactive()
```
