---
title: "Chapter 8"
subtitle: "Data Import"
author: "Aditya Dahiya"
execute: 
  warning: false
  error: false
  message: false
editor_options: 
  chunk_output_type: console
---

# 8.2.4 Exercises

```{r}
#| label: setup

library(tidyverse)
library(gt)
```

### Question 1

**What function would you use to read a file where fields were separated with "\|"?**

Let us first create a dataset with "`|`" delimiter. I used `ChatGPT` to create a random dataset and named it as the data to be imported, i.e., `imp_df` .

I will use the function `read_delim()` to read a file where fields were separated with "\|", as shown below. The output is

```{r}

#| label: tbl-q1-Ex-8.2.4
#| tbl-cap: "Imported data using read_delim() function"

import_df = "Name|Age|Gender|City|Salary
John|28|Male|New York|75000
Emily|22|Female|Los Angeles|60000
Michael|31|Male|Chicago|80000
Jessica|25|Female|Houston|65000
William|29|Male|Miami|70000
Sophia|27|Female|San Francisco|75000
Daniel|24|Male|Seattle|72000
Olivia|30|Female|Boston|78000
James|26|Male|Dallas|67000
Ava|23|Female|Atlanta|62000"

df = read_delim(import_df, delim = "|")

df |>
  gt()
```

### Question 2

**Apart from file, skip, and comment, what other arguments do `read_csv()` and `read_tsv()` have in common?**

While `read_csv()` works for comma separated files, `read_tsv()` works for `TAB` separated files. The arguments for each of them are as follows:-

```{r}
?read_csv
```

### Question 3

**What are the most important arguments to `read_fwf()`?**

### Question 4

**Sometimes strings in a CSV file contain commas. To prevent them from causing problems, they need to be surrounded by a quoting character, like " or '. By default, read_csv() assumes that the quoting character will be ". To read the following text into a data frame, what argument to read_csv() do you need to specify?**

``` r
"x,y\n1,'a,b'"
```

### Question 5

**Identify what is wrong with each of the following inline CSV files. What happens when you run the code?**

``` r
read_csv("a,b\n1,2,3\n4,5,6")

read_csv("a,b,c\n1,2\n1,2,3,4")

read_csv("a,b\n\"1")

read_csv("a,b\n1,2\na,b")

read_csv("a;b\n1;3")
```

### Question 6

**Practice referring to non-syntactic names in the following data frame by:**

a.  **Extracting the variable called `1`.**

b.  **Plotting a scatterplot of `1` vs. `2`.**

c.  **Creating a new column called `3`, which is `2` divided by `1`.**

d.  **Renaming the columns to `one`, `two`, and `three`.**

``` r
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)
```
