---
title: "Chapter 25"
author: "Aditya Dahiya"
subtitle: "Web scraping"
date: 2023-10-06
execute: 
  echo: true
  warning: false
  error: false
  cache: true
  fig-width: 10
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/RfDS2solutions/Chapter25.html"
  description: "Solutions: R for Data Science (2e)"
  twitter: true
  facebook: true
  linkedin: true
  email: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

In the realm of aviation history and military technology, **fighter aircraft** hold a pivotal role, representing the pinnacle of technological prowess and strategic significance. To delve into the comprehensive landscape of fighter aircraft, this assignment embarks on a journey of data acquisition and analysis, leveraging the versatile capabilities of the `R` programming language, specifically the `rvest` package[@rvest] within the `tidyverse` framework[@tidyverse].

Our primary data source, the Wikipedia page titled "[List of Fighter Aircraft](https://en.wikipedia.org/wiki/List_of_fighter_aircraft)," serves as a repository of valuable information encompassing various facets of fighter aircraft production, categorization, and operational status.

This page aims to meticulously extract, process, and analyze the data from this source, shedding light on the countries that have contributed significantly to fighter aircraft production, the prevailing types of aircraft that have dominated the skies, and the intriguing disparities between operational and retired aircraft, including the number of prototypes produced. Through this endeavor, we endeavor to gain profound insights into the historical and contemporary dynamics of fighter aircraft, uncovering patterns and trends that illuminate their pivotal role in military aviation and defense strategies.

Experiment scraping the Wikipedia page of fighter aircraft

```{r}
#| label: setup

# Load libraries
library(tidyverse)
library(rvest)
library(gt)
library(gtExtras)
library(scales)
```

```{r}
#| echo: false
#| eval: true


df <- read_csv("docs/temp_ch25.csv")

```

```{r}
#| label: web-scraping-dont-evaluate
#| eval: false
#| echo: true

# The web-address to harvest
url <- "https://en.wikipedia.org/wiki/List_of_fighter_aircraft"

# Read in the html file from the URL
url_html <- read_html(url)

# Create a nice looking tibble from the data
df <- url_html |>
  html_element("table") |>
  html_table() |>
  
  # house-keeping tasks: cleaning names etc.
  janitor::clean_names() |>
  rename(
    numbers = no,
    year = date
  ) |>
  select(-notes) |>
  mutate(
    numbers = parse_number(numbers),
    class = str_replace_all(class, 
                            pattern = "\"", 
                            replacement = ""),
    class = na_if(class, ""),
    status = fct(status, levels = c("Abandoned",
                                    "Prototype",
                                    "In Development",
                                    "Retired",
                                    "Operational"))
  )
```

Displaying the complete data, in a nicely formatted table at @fig-complete-data : --

```{r}
#| label: fig-complete-data
#| fig-cap: "The complete data on fighter aircraft produced by different countries"

df |>
  gt() |>
  cols_label_with(fn = ~ janitor::make_clean_names(., "title")) |>
  opt_interactive(page_size_default = 5,
                  use_highlight = TRUE,
                  use_resizers = TRUE)
```

Using the data for some analysis

In order to visually convey the the total number of fighter aircraft manufactured by each nation, we employ the powerful visualization tool provided by [Datawrapper](https://www.datawrapper.de/ "https://www.datawrapper.de/"), and create a pie chart: --

```{=html}
<iframe title="4 countries account for more than 4/5ths of all fighter aircraft ever made!" aria-label="Pie Chart" id="datawrapper-chart-RABvd" src="https://datawrapper.dwcdn.net/RABvd/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="501" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r=0;r<e.length;r++)if(e[r].contentWindow===a.source){var i=a.data["datawrapper-height"][t]+"px";e[r].style.height=i}}}))}();
</script>
```
Now, we will visualize the production of fighter aircraft over different decades, categorized by their design years and attributed to various countries, using a stacked bar chart in @fig-stacked-bar .

```{r}
#| label: fig-stacked-bar
#| fig-cap: "A stacked bar chart for country-wise aircraft made in each decade"

# A stacked bar chart for country-wise aircraft made in each decade

df |>
  mutate(
    country = fct(country),
    decade = fct(paste0(floor(year / 10) * 10, "s"), 
                 levels = paste0(seq(1910, 2020, by = 10), "s"))
  ) |>
  group_by(decade, country) |>
  count(wt = numbers, 
        name = "nos_planes") |>
  
  # Ungroup so that factor lumping works
  ungroup() |>
  
  # Remove 2020s as very few observations; and irrelevant to the message
  filter(decade != "2020s") |>
  
  # Lump into 5 factors for easy display of coloured stacked bar chart
  mutate(country = fct_lump_n(country, n = 5, w = nos_planes)) |> 
  
  # Start plotting
  ggplot(aes(x = decade,
             y = nos_planes,
             fill = country,
             label = country,
             text = paste0("Aircrafts produced in the decade: ", nos_planes))
         ) +
  geom_bar(stat = "identity",
           position = "stack") +
  scale_y_continuous(labels = label_number_si()) +
  labs(x = NULL, 
       y = "Number of aircraft",
       title = "The World War decades saw maximum aircraft production",
       subtitle = "Germany dominated the prodcution in the pre WW-II 1930s\nBut, therafter, US and USSR dominated production right upto the end of cold war in 1980s",
       fill = NULL) +
  scale_fill_manual(values = c("#bbbfc9",
                               "red",
                               "#8491b5",
                               "#5c6580",
                               "#576cad",
                               "#f5f114")) +
  guides(fill = guide_legend(nrow = 1)) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.grid = element_blank(),
        plot.title.position = "plot")

```

Here in @fig-operational-planes are the most produced operational aircraft still operational as of today.

```{r}
#| label: fig-operational-planes
#| fig-cap: "Most produced, and currently operational aircraft"


df |>
  filter(status == "Operational") |>
  arrange(desc(numbers)) |>
  slice_head(n = 10) |>
  mutate(type = if_else(type == "McDonnell Douglas F/A-18 Hornet",
                        "F/A-18 Hornet",
                        type)) |>
  ggplot(aes(x = numbers,
             y = reorder(type, numbers),
             fill = country)) +
  geom_bar(stat = "identity") +
  theme_void() +
  geom_text(aes(x = 0,
                label = type),
            hjust = "left") +
  geom_text(aes(label = numbers),
            hjust = -0.3) +
  labs(fill = NULL,
       title = "The most produced, and still operational, fighter planes belong to US or USSR",
       subtitle = "MiG-21 is most produced craft, outnumbering the next highest by more than twice!")  +
  scale_fill_manual(values = c("#f5f584", "lightblue", "#828afa", "#f28374")) +
  theme(plot.title.position = "panel",
        legend.position = "bottom")
```

But, if we look at the most mass-produced ever made, all of them are retired as shown below in @fig-retired-planes .

```{r}
#| label: fig-retired-planes
#| fig-cap: "Most produced aircraft of all time: all of them are retired"

df |>
  arrange(desc(numbers)) |>
  slice_head(n = 10) |>
  ggplot(aes(x = numbers,
             y = reorder(type, numbers),
             fill = country)) +
  geom_bar(stat = "identity") +
  theme_void() +
  geom_text(aes(x = 0,
                label = type),
            hjust = "left") +
  geom_text(aes(label = scales::comma(numbers)),
            hjust = -0.3) +
  labs(fill = NULL,
       title = "However, the most produced fighter planes ever, are all retired!",
       subtitle = "Historically, Germany and UK, mass-produced the planes they no longer use.")  +
  scale_fill_manual(values = c("#97979c", "#54b8ff", "#828afa", "#f28374")) +
  theme(plot.title.position = "panel",
        legend.position = "bottom")
```

Now, looking at the **class** of aircraft, the most produced type of aircraft, all time are shown in @fig-type-all-time .

```{r}
#| label: fig-type-all-time
#| fig-cap: "Number of aircraft, ever produced, by the type or class of aircraft"


df |>
  mutate(class = if_else(class == "fighter-bomber",
                         "Fighter-bomber",
                         class)) |>
  count(class, sort = TRUE, wt = numbers) |>
  drop_na() |>
  mutate(
    class = fct(class),
    class = fct_lump_n(class, n = 9, w = n)
  ) |>
  count(class, sort = TRUE, wt = n) |>
  ggplot(aes(x = n,
             y = fct_rev(class))) +
  geom_bar(stat = "identity", fill = "grey") +
  theme_void() +
  geom_text(aes(x = 0,
                label = class),
            hjust = "left") +
  geom_text(aes(label = scales::comma(n)),
            hjust = -0.1) +
  labs(fill = NULL,
       title = "Fighter-bombers were the most common type of aircraft ever produced!") +
  theme(plot.title.position = "panel",
        legend.position = "bottom")
```

Laslty, the @tbl-prototypes provides the top 10 fighter aircraft types that predominantly remained in the prototype stage, offering valuable insights into classes which remained experimental only in military aviation development.

```{r}
#| label: tbl-prototypes
#| tbl-cap: "Aircraft which had highest percentage of total produced planes remaining as prototypes only"

df |>
  group_by(class, status) |>
  count(wt = numbers) |>
  group_by(class) |>
  mutate(total = sum(n)) |>
  filter(status == "Prototype") |>
  mutate(prop = n/total) |>
  filter(total >= 10) |>
  arrange(desc(prop)) |>
  drop_na() |>
  select(-status) |>
  ungroup() |>
  rename(
    type_of_aircraft = class,
    number_of_prototypes = n,
    total_aircraft_produced = total,
    percentage_of_prototypes = prop
  ) |>
  slice_head(n = 10) |>
  gt() |>
  cols_label_with(fn = ~ janitor::make_clean_names(., "title")) |>
  fmt_percent(columns = percentage_of_prototypes, decimals = 1) |>
  gt_theme_538()
```
