---
title: "Chapter 19"
author: "Aditya Dahiya"
subtitle: "Missing Values"
date: 2023-10-02
execute: 
  echo: true
  warning: false
  error: false
  cache: true
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/RfDS2solutions/Chapter19.html"
  description: "Solutions: R for Data Science (2e)"
  twitter: true
  facebook: true
  linkedin: true
  email: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r}
#| label: setup

library(tidyverse)
library(janitor)
library(gt)
library(gtExtras)
```

# **19.3.4 Exercises**

## Question 1

**Can you find any relationship between the carrier and the rows that appear to be missing from `planes`?**

Yes, as we can see in the @tbl-q1-ex3 , the airline carriers `MQ` and `AA` have most of their aircrafts' tail numbers missing from the `planes` data-set, apart from few other carriers that have a small percentage of their data missing.

```{r}
#| label: tbl-q1-ex3
#| tbl-cap: "Percentage of tail numbers of each carrier that is missing from planes data set"
#| tbl-cap-location: top

library(nycflights13)
data("flights")
data("planes")

# Create a vector of carriers that have tailsnums missing in planes
car_vec = flights |>
  distinct(tailnum, carrier) |>
  anti_join(planes) |>
  distinct(carrier) |>
  as_vector() |>
  unname()

# Find total tailnums for these carriers
total_tails = flights |>
  filter(carrier %in% car_vec) |>
  group_by(carrier) |>
  summarize(
    total_aircrafts = n_distinct(tailnum)
    )

flights |>
  distinct(tailnum, carrier) |>
  anti_join(planes) |>
  count(carrier, name = "missing_tailnums") |>
  full_join(total_tails) |>
  mutate(percentage_missing = missing_tailnums/total_aircrafts) |>
  arrange(desc(percentage_missing)) |>
  gt() |>
  gt_theme_538() |>
  cols_label_with(fn = ~ janitor::make_clean_names(., case = "title")) |>
  fmt_percent(columns = percentage_missing) |>
   tab_style(
    style = list(cell_text(weight = "bold") ),
    locations = cells_body(columns = percentage_missing)) 
```

# Other practice stuff

Since this chapter has only one question, I will also explore another package `naniar`[@naniar] and `visdat`[@visdat]that deals with missing values. Some key learning: --

```{r}
#| label: loading datasets and packages
library(naniar)
library(visdat)
library(nycflights13)
data("flights")
```

1.  **Visualize a data.frame to see what it contains**

    ```{r}
    planes |> 
      vis_dat() +
      labs(title = "A vis_dat() output to see the contents of a data-frame")

    flights |>
      slice_sample(n = 100) |>
      vis_dat() +
      labs(title = "A vis_dat() output to see the contents of a data-frame",
           col = "Type of variable") +
      theme(legend.position = "bottom") +
      scale_fill_brewer(palette = "Accent")
    ```

2.  Show missingness in particular, with percentages: --

    ```{r}
    flights |>
      slice_sample(n = 100) |>
      vis_miss() +
      labs(title = "A vis_miss() output to see the missing values",
           col = "Type of variable")
    ```

3.  **Show the proportion of missingness for each variable**

    ```{r}
    flights |>
      gg_miss_var(show_pct = TRUE) +
      labs(title = "Percentage missing values for each variable in flights dataset",
           y = NULL, 
           x = "Percentage missing values (%)")
    ```

4.  To replace a value, say "99" or "-99" with `NA`, we use `naniar::replace_with_na()` or `dplyr::na_if()` . To replace `NA` with a given value we use, `dplyr::replace_na()`

5.  **Using `vis_expect()` to see values that fulfill certain conditions**

    ```{r}
    flights |>
      select(day, dep_delay, arr_delay, air_time) |>
      slice_sample(n = 200) |>
      vis_expect(~.x >= 10)
    ```
