---
title: "Chapter 17"
author: "Aditya Dahiya"
subtitle: "Factors"
date: 2023-09-14
execute: 
  echo: true
  warning: false
  error: false
  cache: true
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/RfDS2solutions/Chapter17.html"
  description: "Solutions: R for Data Science (2e)"
  twitter: true
  facebook: true
  linkedin: true
  email: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r}
#| label: setup

library(tidyverse)
library(ggthemes)
library(gt)
library(gtExtras)
data("gss_cat")
```

**Important Points**

-   Most rename and recoding functions in the tidyverse, the new values go on the left and the old values go on the right.

# **17.3.1 Exercise**

## Question 1

**Explore the distribution of `rincome` (reported income). What makes the default bar chart hard to understand? How could you improve the plot?**

The default bar chart is hard to understand because: --

1.  It is vertical, and the names of categories overlap on x-axis.

2.  The "Not applicable" category is before the lowest income group. Thus, the pattern is disturbed.

We could improve the plot, as shown in @fig-q1-ex3, by: ---

1.  Making it into a horizontal bar chart to allow space and easy reading of categories of income levels.

2.  Move the "Not Applicable" level after the highest income level, along-side "Refused", "Dont' know" and "No answer".

3.  Further, we could remove non-data ink, as per principles of Mr. Tufte to make our pattern stand out. Also, we could create a separate colouring scheme for data outside the income levels.

```{r}
#| label: fig-q1-ex3
#| fig-cap: "Improved bar chart"

no_levels = levels(gss_cat$rincome)[c(1:3, 16)]

gss_cat |>
  mutate(col_level = rincome %in% no_levels) |>
  ggplot(aes(y = fct_relevel(rincome, 
                             "Not applicable",
                             after = 3),
             fill = col_level)) +
  geom_bar() +
  theme_minimal() +
  labs(x = "Number of respondents", y = NULL,
       title = "Income Levels of respondents in General Social Survey") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  scale_fill_manual(values = c("#3d3b3b", "#999494"))
```

## Question 2

**What is the most common `relig` in this survey? What's the most common `partyid`?**

The most common `relig` is "Protestant". And, the most common `partyid` is "Independent".

```{r}
gss_cat |>
  count(relig, sort = TRUE)

gss_cat |>
  count(partyid, sort = TRUE)
```

## Question 3

**Which `relig` does `denom` (denomination) apply to? How can you find out with a table? How can you find out with a visualization?**

We can see from the code below that more than one factor values in `denom` (denomination) occur only in "Protestant", "Christian" and "Other" religions. To explore further, we can cross-tabulate religion and denomination, as shown in @tbl-q3-ex3, and realize that the only religion to which denomination really applies to is "Protestant".

We could also do a visualization as in @fig-q3a-ex3 .

```{r}
gss_cat |>
  group_by(relig) |>
  summarise(n = n_distinct(denom)) |>
  arrange(desc(n)) |>
  filter(n > 1)
```

```{r}
#| label: tbl-q3-ex3
#| tbl-cap: "Cross-Table of the deominations within the three religions"

gss_cat |>
  filter(relig %in% c("Protestant", "Christian", "Other")) |>
  group_by(relig, denom) |>
  tally() |>
  spread(relig, n) |>
  arrange(desc(Christian)) |>
  gt() |>
  gt_theme_538()
```

```{r}
#| label: fig-q3a-ex3
#| tbl-cap: "Visualization of the number of deominations within religions"

gss_cat |>
  group_by(relig) |>
  summarise(n = n_distinct(denom)) |>
  arrange(desc(n)) |>
  ggplot(aes(y = reorder(relig, n), x = n)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(x = "Number of denominations", y = NULL,
       title = "Only Protestant religion has demoninations within it")
```

# **17.4.1 Exercises**

## Question 1

**There are some suspiciously high numbers in `tvhours`. Is the mean a good summary?**

No, mean is not a good summary as the distribution of `tvhours` is right skewed. Instead, we should use median as a summary measure.

```{r}
gss_cat |>
  drop_na() |>
  mutate(tvhours = as_factor(tvhours)) |>
  ggplot(aes(x = tvhours)) +
  geom_bar(col = "black", fill = "white") +
  theme_clean() +
  labs(x = "Hours per day spent watching TV",
       y = "Numbers", title = "Distribution of TV Hours is right skewed")
```

## Question 2

**For each factor in `gss_cat` identify whether the order of the levels is arbitrary or principled.**

The variables in the `gss_cat` data-set which are factors are: ---

| Factor Variable | Levels                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                           | Order is arbitrary or principled                                                                                 |
|------------|-------------------------------------------------|------------|
| `marital`       | No answer, Never married, Separated, Divorced, Widowed and, Married                                                                                                                                                                                                                                                                                                                                                                                                                                                                              | Arbitrary, since they are not in a specific order                                                                |
| `race`          | Other, Black, White and, Not applicable                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                          | Arbitrary, since they are not in a specific order                                                                |
| `rincome`       | No answer, Don't know, Refused, \$25000 or more, \$20000 - 24999, \$15000 - 19999, \$10000 - 14999, \$8000 to 9999, \$7000 to 7999, \$6000 to 6999, \$5000 to 5999, \$4000 to 4999, \$3000 to 3999, \$1000 to 2999, Lt \$1000 and, Not applicable                                                                                                                                                                                                                                                                                                | Principled, since the income levels are in a specified increasing or decreasing order, with few levels arbitrary |
| `partyid`       | No answer, Don't know, Other party, Strong republican, Not str republican, Ind,near rep, Independent, Ind,near dem, Not str democrat and, Strong democrat                                                                                                                                                                                                                                                                                                                                                                                        | Partly Principled, as there are two extremes, and then levels in the middle.                                     |
| `relig`         | No answer, Don't know, Inter-nondenominational, Native american, Christian, Orthodox-christian, Moslem/islam, Other eastern, Hinduism, Buddhism, Other, None, Jewish, Catholic, Protestant and, Not applicable                                                                                                                                                                                                                                                                                                                                   | Arbitrary, as the religions are not in a specific order.                                                         |
| `denom`         | No answer, Don't know, No denomination, Other, Episcopal, Presbyterian-dk wh, Presbyterian, merged, Other presbyterian, United pres ch in us, Presbyterian c in us, Lutheran-dk which, Evangelical luth, Other lutheran, Wi evan luth synod, Lutheran-mo synod, Luth ch in america, Am lutheran, Methodist-dk which, Other methodist, United methodist, Afr meth ep zion, Afr meth episcopal, Baptist-dk which, Other baptists, Southern baptist, Nat bapt conv usa, Nat bapt conv of am, Am bapt ch in usa, Am baptist asso and, Not applicable | Arbitrary, as the denominations are not in a specific order.                                                     |

## Question 3

**Why did moving "Not applicable" to the front of the levels move it to the bottom of the plot?**

Moving "Not applicable" to the front of the levels move it to the bottom of the plot, because `ggplot2` plots the levels in increasing order, starting bottom's upwards.

# **17.5.1 Exercises**

## Question 1

**How have the proportions of people identifying as Democrat, Republican, and Independent changed over time?**

As reflected in the @fig-q1-ex5, the proportions of people identifying as Democrat has slightly increased, Republican has slightly decreased, and Independent has increased, over the period of 15 years reflected in the data-set.

```{r}
#| label: fig-q1-ex5
#| fig-cap: "Stacked bar chart of the partyid in the data-set"


gss_cat |>
  mutate(
    partyid = fct_collapse(partyid,
      "Republican"  = c("Strong republican",  "Not str republican"),
      "Democrat"    = c("Strong democrat", "Not str democrat"),
      "Independent" = c("Independent", "Ind,near dem", "Ind,near rep"),
      "Others"      = c("No answer", "Don't know", "Other party")
    )
  ) |>
  group_by(year, partyid) |>
  count() |>
  ggplot(aes(x = year, y = n, fill = partyid)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = c("lightgrey", "red", "grey", "blue")) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = "Year", y = "Proportion of respondents", fill = "Party",
       subtitle = "Proportion of republicans has decreased, while that of independents has increased over the years",
       title = "In 15 years, share of parties' supporters has changed")
```

## Question 2

**How could you collapse `rincome` into a small set of categories?**

We could collapse the `rincome` into a small set of categories using the following functions: --

-   `fct_lump_n()`

-   `fct_lump_lowfreq()`

-   `fct_lump_min()`

-   `fct_lump_prop()`

-   `fct_lump()`

-   `fct_collapse()`

```{r}
gss_cat|>
  mutate(
    rincome = fct_lump_n(rincome, n = 6)
  ) |>
  group_by(rincome) |>
  count() |>
  arrange(desc(n)) |>
  ungroup() |>
  gt() |>
  cols_label(rincome = "Annual Income",
             n = "Numbers")
```

## Question 3

**Notice there are 9 groups (excluding other) in the `fct_lump` example above. Why not 10? (Hint: type [`?fct_lump`](https://forcats.tidyverse.org/reference/fct_lump.html), and find the default for the argument `other_level` is "Other".)**

Yes, there are 9 groups (excluding other) in this example, as shown below also in @fig-q3-ex5. This is because `n = 10` argument limits the total groups to 10, and the function needs one group for ***"Other"***, i.e. all other groups whose count is lesser than top 9 groups. Thus, the groups shown are 9, with 1 as ***"Other"*** (at the end).

```{r}
#| label: fig-q3-ex5
#| fig-cap: "Table of number of respondents from each of top 10 religions, including Other"

gss_cat |>
  mutate(relig = fct_lump_n(relig, n = 10)) |>
  count(relig) |>
  gt()
```
