---
title: "Chapter 26"
author: "Aditya Dahiya"
subtitle: "Functions"
date: 2023-10-09
execute: 
  echo: true
  warning: false
  error: false
  cache: true
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/RfDS2solutions/Chapter26.html"
  description: "Solutions: R for Data Science (2e)"
  twitter: true
  facebook: true
  linkedin: true
  email: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r}
#| label: setup

library(tidyverse)
library(gt)
library(gtExtras)
library(nycflights13)
data("flights")
```

Important stuff from Chapter 26, [R for Data Science (2e)](https://r4ds.hadley.nz/functions)

-   To find the definition of a function that you've written, place the cursor on the name of the function and press `F2`.

-   To quickly jump to a function, press `Ctrl + .` to open the fuzzy file and function finder and type the first few letters of your function name. You can also navigate to files, Quarto sections, and more, making it a very handy navigation tool.

-   In R, the **`{{}}`** (double curly braces) operator is used for ***unquoting arguments in functions***.

-   The **`:=`** operator is used for ***in-place modification or assignment of columns*** within a data.table object. It is typically used to add or update columns by reference, meaning it doesn't create a new `data.table` but instead modifies the existing one directly. This can be very efficient for large datasets because it avoids unnecessary copying of data.

# **26.2.5 Exercises**

## Question 1

**Practice turning the following code snippets into functions. Think about what each function does. What would you call it? How many arguments does it need?**

```         
mean(is.na(x))
mean(is.na(y))
mean(is.na(z))
```

The following function computes the proportion of missing values in a vector. It needs one argument.

```{r}
prop_missing <- function(x){
  mean(is.na(x))
}
```

\_\_\_\_\_\_\_\_\_\_\_\_\_

```         
x / sum(x, na.rm = TRUE)
y / sum(y, na.rm = TRUE)
z / sum(z, na.rm = TRUE)
```

The following function computes the proportion of each element of a vector to the sum of the vector. It needs one argument.

```{r}
prop_element <- function(x){
  x / sum(x, na.rm = TRUE)
}
```

\_\_\_\_\_\_\_\_\_\_\_\_\_

```         
round(x / sum(x, na.rm = TRUE) * 100, 1)
round(y / sum(y, na.rm = TRUE) * 100, 1)
round(z / sum(z, na.rm = TRUE) * 100, 1)
```

This function below computes the percentage of each value (as compared to the sum of the vector of values) within one decimal place.

```{r}
perc_element <- function(x){
  round(x / sum(x, na.rm = TRUE) * 100, 1)
}
```

## Question 2

**In the second variant of `rescale01()`, infinite values are left unchanged. Can you rewrite `rescale01()` so that `-Inf` is mapped to 0, and `Inf` is mapped to 1?**

```{r}
rescale01 <- function(x) {
  
  # Replace -Inf with the minimum number (other than -Inf)
  min_value <- min(x[is.finite(x)])
  x[x == -Inf] <- min_value

  # Replace +Inf with the maximum number (other than +Inf)
  max_value <- max(x[is.finite(x)])
  x[x == Inf] <- max_value

  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
```

## Question 3

**Given a vector of birth-dates, write a function to compute the age in years.**

```{r}
# Function to compute age in years from birth dates

compute_age <- function(birth_dates) {
  
  # Convert the birth dates to Date objects
  birth_dates <- as.Date(birth_dates)
  
  # Calculate the current date
  current_date <- Sys.Date()
  
  # Calculate age in years using lubridate
  ages <- interval(birth_dates, current_date) %/% years(1)
  
  # Return the ages as a numeric vector
  return(ages)
}
```

## Question 4

**Write your own functions to compute the variance and skewness of a numeric vector. You can look up the definitions on Wikipedia or elsewhere.**

To compute the variance and skewness of a numeric vector, we can define custom functions for each: --

1.  **Variance Calculation** using the formula:

``` r
# Function to compute the variance
compute_variance <- function(x) {
  n <- length(x)
  x_bar <- mean(x)
  sum_squared_diff <- sum((x - x_bar)^2)
  variance <- sum_squared_diff / (n - 1)
  return(variance)
}
```

2.  **Skewness Calculation** using the formula:

``` r
# Function to compute the skewness
compute_skewness <- function(x) {
  n <- length(x)
  x_bar <- mean(x)
  std_dev <- sqrt(compute_variance(x))  # Using previously defined variance function
  skewness <- sum((x - x_bar)^3) / ((n - 1) * std_dev^3)
  return(skewness)
}
```

These custom functions `compute_variance` and `compute_skewness` will calculate the variance and skewness of a numeric vector based on the provided formulas, without relying on external packages.

## Question 5

**Write `both_na()`, a summary function that takes two vectors of the same length and returns the number of positions that have an `NA` in both vectors.**

```{r}

# Method 1
both_na <- function(x,y){
  
  x_na <- which(is.na(x))
  y_na <- which(is.na(y))
  
  # values of x which are also present in y
  common <- x_na %in% y_na
  return(common)
}

# Method 2
both_na <- function(vector1, vector2) {
  if (length(vector1) != length(vector2)) {
    stop("Both vectors must have the same length.")
  }
  
  # Find the indices where both vectors have NA values
  na_indices <- which(is.na(vector1) & is.na(vector2))
  
  return(na_indices)
}
```

## Question 6

**Read the documentation to figure out what the following functions do. Why are they useful even though they are so short?**

```         
is_directory <- function(x) {
  file.info(x)$isdir
 }
```

This function tells whether an object is a file or a directory. It is useful because it will return a logical value, i.e., either `TRUE` or `FALSE` and thus can be used inside other operations, rather than using the `$` operator; and is easy to remember in English.

\_\_\_\_\_\_\_\_\_

```         
is_readable <- function(x) {
  file.access(x, 4) == 0
}
```

This function tells us whether the file has a read permission or not, i.e, whether there is access to the file or not. It is useful because one does not have to remember the `mode` argument values for `file.access()` function, and is easy to remember in English for further use in `for` loops etc.

# **26.3.5 Exercises**

## Question 1

**Using the datasets from `nycflights13`, write a function that:**

```{r}
# Creating a function to display the results in a nice way

display_results <- function(data){
  data |>
  slice_head(n = 5) |> 
  gt() |> 
  cols_label_with(fn = ~ janitor::make_clean_names(., "title")) |>
  gt_theme_538()
}

```

1.  **Finds all flights that were cancelled (i.e. `is.na(arr_time)`) or delayed by more than an hour.**

    ```         
    flights |> filter_severe()
    ```

    ```{r}
    filter_severe <- function(data, arrival_delay) {
      data |>
        filter(is.na({{arrival_delay}}) | ({{arrival_delay}} > 60))
    }

    # Runinng an example to show it works
    flights |>
      filter_severe(arr_delay) |>
      select(1:10) |>
      display_results()
    ```

2.  **Counts the number of cancelled flights and the number of flights delayed by more than an hour.**

    ```         
    flights |> group_by(dest) |> summarize_severe()
    ```

    ```{r}
    summarize_severe <- function(data, arrival_delay){
      data |>
        summarize(
          cancelled_flights = sum(is.na({{arrival_delay}})),
          delayed_flights = sum({{arrival_delay}} > 60, na.rm = TRUE)
        )
    }

    # Runinng an example to show it works
    flights |>
      group_by(dest) |>
      summarize_severe(arr_delay) |>
      display_results()
    ```

3.  **Finds all flights that were cancelled or delayed by more than a user supplied number of hours:**

    ```         
    flights |> filter_severe(hours = 2)
    ```

    ```{r}
    filter_severe <- function(data, hours){
      data |>
        filter(arr_delay > (hours*60))
    }

    # Running an example to show that it works
    flights |>
      filter_severe(hours = 4) |>
      select(1:10) |>
      display_results()
    ```

4.  **Summarizes the weather to compute the minimum, mean, and maximum, of a user supplied variable:**

    ```         
    weather |> summarize_weather(temp)
    ```

    ```{r}
    summarize_weather <- function(data, variable){
      data |>
        summarize(
          mean = mean({{variable}}, na.rm = TRUE),
          minimum = min({{variable}}, na.rm = TRUE),
          maximum = max({{variable}}, na.rm = TRUE)
        )
    }
    # Runinng an example to show it works
    weather |>
      group_by(origin) |>
      summarize_weather(temp) |>
      display_results() |> fmt_number(decimals = 2)
    ```

5.  **Converts the user supplied variable that uses clock time (e.g., `dep_time`, `arr_time`, etc.) into a decimal time (i.e. hours + (minutes / 60)).**

```         
flights |> standardize_time(sched_dep_time)
```

```{r}
# Method 1
standardize_time <- function(data, variable){
  data |>
    mutate(std_time = round({{variable}} %/% 100) + (({{variable}} %% 100)/60), 2) |>
    relocate(std_time, .after = {{variable}})
}

# Method 2 (after learning use of ":=" in Section 26.4.2)
standardize_time <- function(data, variable){
  data |>
    mutate({{variable}} := round({{variable}} %/% 100) + (({{variable}} %% 100)/60), 2) 
}

# Runinng an example to show it works
flights |>
  standardize_time(sched_dep_time) |>
  select(1:10) |>
  display_results()


```

## Question 2

**For each of the following functions list all arguments that use tidy evaluation and describe whether they use data-masking or tidy-selection: [`distinct()`](https://dplyr.tidyverse.org/reference/distinct.html), [`count()`](https://dplyr.tidyverse.org/reference/count.html), [`group_by()`](https://dplyr.tidyverse.org/reference/group_by.html), [`rename_with()`](https://dplyr.tidyverse.org/reference/rename.html), [`slice_min()`](https://dplyr.tidyverse.org/reference/slice.html), [`slice_sample()`](https://dplyr.tidyverse.org/reference/slice.html).**

In the context of the `dplyr` package in `R`, ***tidy evaluation*** refers to a method of non-standard evaluation where you can use variables as arguments to functions and refer to those variables within the context of a data frame. It has two parts: --

1.  **Data Masking**: Data masking involves creating a temporary environment within a function where external variables from the global environment are hidden or masked. This means that variables from the global environment are temporarily unavailable inside the function, preventing unintended interactions or conflicts. For example:

    ```{r}
    data <- data.frame(x = 1:5, y = 6:10)

    custom_function <- function(df) {
      
      # Using data masking to prevent access to global 'data' variable
      df %>%
        mutate(z = x + y)
    }

    custom_function(data)
    ```

    In this example, the `custom_function` takes a data frame `df` as an argument and uses the `mutate()` function from dplyr to create a new variable `z`. Inside the function, data masking ensures that `x` and `y` refer to the columns within the `df` argument, not the global `data` data frame.

2.  **Tidy Selection**: Tidy selection involves selecting or manipulating columns within a data frame using non-standard evaluation (NSE). It allows you to refer to column names as symbols or expressions, facilitating dynamic and programmatic column selection. For example: --

    ```{r}
    data <- data.frame(x = 1:5, y = 6:10)

    selected_columns <- c("x", "y")

    data %>%
      select(all_of(selected_columns))
    ```

    In this example, the `select()` function uses tidy selection to select columns specified in the `selected_columns` vector.

Thus, Data masking primarily deals with the environment of the function, while tidy selection focuses on column selection and manipulation within a data frame.

All of the following functions use ***Tidy Evaluation*** in the following way: --

+------------------+------------------------------------------------------------------+----------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Function         | Description                                                      | Tidy Selection | Arguments                                                                                                                                                                                                                           |
+==================+==================================================================+================+=====================================================================================================================================================================================================================================+
| `distinct()`     | Select distinct rows based on specified columns                  | Yes            | -   The **`data`** argument does not involve Data Masking or Tidy Selection.                                                                                                                                                        |
|                  |                                                                  |                |                                                                                                                                                                                                                                     |
|                  |                                                                  |                | -   The **`...`** argument supports Tidy Selection methods for specifying columns. It also employs data masking to prevent global environment variables from interfering.                                                           |
|                  |                                                                  |                |                                                                                                                                                                                                                                     |
|                  |                                                                  |                | -   The **`.keep_all`** argument does not involve Data Masking or Tidy Selection.                                                                                                                                                   |
+------------------+------------------------------------------------------------------+----------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `count()`        | Count the number of rows in groups defined by specific variables | Yes            | -   The **`data`** argument does not involve Data Masking or Tidy Selection.                                                                                                                                                        |
|                  |                                                                  |                |                                                                                                                                                                                                                                     |
|                  |                                                                  |                | -   The **`...`** argument supports Tidy Selection methods for specifying columns. It also employs data masking to prevent global environment variables from interfering.                                                           |
|                  |                                                                  |                |                                                                                                                                                                                                                                     |
|                  |                                                                  |                | -   The **`.name`** argument does not involve Data Masking or Tidy Selection.                                                                                                                                                       |
+------------------+------------------------------------------------------------------+----------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `group_by()`     | Group data by specific columns                                   | Yes            | -   The **`data`** argument does not involve Data Masking or Tidy Selection.                                                                                                                                                        |
|                  |                                                                  |                |                                                                                                                                                                                                                                     |
|                  |                                                                  |                | -   The **`...`** argument supports data masking to prevent global environment variables from interfering. It does not use tidy selection.                                                                                          |
|                  |                                                                  |                |                                                                                                                                                                                                                                     |
|                  |                                                                  |                | -   The **`.add`** argument does not involve Data Masking or Tidy Selection.                                                                                                                                                        |
+------------------+------------------------------------------------------------------+----------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `rename_with()`  | Rename columns based on a function or expression                 | Yes            | -   The **`data`** argument does not involve Data Masking or Tidy Selection.                                                                                                                                                        |
|                  |                                                                  |                |                                                                                                                                                                                                                                     |
|                  |                                                                  |                | -   The **`.fn`** and **`...`** arguments support Tidy Selection methods for specifying columns and renaming logic.                                                                                                                 |
|                  |                                                                  |                |                                                                                                                                                                                                                                     |
|                  |                                                                  |                | -   The **`.cols`** argument supports Tidy Selection methods for specifying the columns to apply renaming to. It also employs data masking to prevent global environment variables from interfering.                                |
+------------------+------------------------------------------------------------------+----------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `slice_min()`    | Filter rows corresponding to the minimum value of a column       | Yes            | -   The **`data`** and **`n`** arguments do not involve Data Masking or Tidy Selection.                                                                                                                                             |
|                  |                                                                  |                |                                                                                                                                                                                                                                     |
|                  |                                                                  |                | -   The **`order_by`** argument supports Tidy Selection methods for specifying the column to determine the order for finding minimum values. It also employs data masking to prevent global environment variables from interfering. |
+------------------+------------------------------------------------------------------+----------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `slice_sample()` | Randomly sample rows from a data frame                           | Yes            | -   The **`data`**, **`n`**, **`replace`**, and **`.groups`** arguments do not involve Data Masking or Tidy Selection.                                                                                                              |
|                  |                                                                  |                |                                                                                                                                                                                                                                     |
|                  |                                                                  |                | -   The **`weight`** argument supports Tidy Selection methods for specifying the sampling weight variable.                                                                                                                          |
+------------------+------------------------------------------------------------------+----------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

In summary, all of the functions mentioned above support tidy evaluation for specifying arguments related to column selection or manipulation.

## Question 3

**Generalize the following function so that you can supply any number of variables to count.**

```         
count_prop <- function(df, var, sort = FALSE) {
  df |>
    count({{ var }}, sort = sort) |>
    mutate(prop = n / sum(n))
}
```

Here is the modified function to allow user to supply any number of variables to count: --

```{r}
count_prop <- function(df, vars, sort = FALSE) {
  df |>
    count(pick({{ vars }}), sort = sort) |>
    mutate(prop = n / sum(n))
}

# Testing the results
flights |>
  count_prop(c(dest, origin)) |>
  slice_head(n = 10) |> 
  gt() |> gt_theme_538() |> fmt_number(prop, decimals = 4)
  
```

# **26.4.4 Exercises**

**Build up a rich plotting function by incrementally implementing each of the steps below:**

## Question 1

**Draw a scatter-plot given dataset and `x` and `y` variables.**

```{r}
scatterplot <- function(data, x, y){
  data |>
    ggplot(aes(x = {{x}},
               y = {{y}})) +
    geom_point() +
    theme_classic()
}
```

## Question 2

**Add a line of best fit (i.e. a linear model with no standard errors).**

```{r}
scatterplot <- function(data, x, y){
  data |>
    ggplot(aes(x = {{x}},
               y = {{y}})) +
    geom_point() +
    geom_smooth(method = "lm",
                formula = {{y}} ~ {{x}},
                se = FALSE) +
    theme_classic()
}
```

## Question 3

**Add a title.**

```{r}
scatterplot <- function(data, x, y){
    ggplot(data, aes(x = {{x}},
               y = {{y}})) +
    geom_point() +
    geom_smooth(method = "lm",
                formula = {{y}} ~ {{x}},
                se = FALSE) +
    labs(title = rlang::englue("A scatter plot of {{y}} vs. {{x}}")) +
    theme_classic()
}
```

# **26.5.1 Exercises**

## Question 1

**Read the source code for each of the following two functions, puzzle out what they do, and then brainstorm better names.**

**`is_prefix()`**

```         
# is_prefix
f1 <- function(string, prefix) {
  str_sub(string, 1, str_length(prefix)) == prefix
}
```

**`match_length()`**

```         
f3 <- function(x, y) {
  rep(y, length.out = length(x))
}
```

## Question 2

**Take a function that you've written recently and spend 5 minutes brainstorming a better name for it and its arguments.**

Here are some suggestions for renaming **`html_nodes()`** and its arguments (from the package `rvest`):

1.  Original Function: **`html_nodes()`**

    -   **`html_extract()`** - This name suggests the action of extracting nodes from an HTML document.

    -   **`select_html_nodes()`** - A more descriptive name that clarifies the purpose of the function.

2.  Argument **`css`**:

    -   **`selector`** - A more explicit name to indicate that this argument takes CSS selectors.

    -   **`element_selector`** - If the purpose of the argument is to select HTML elements, this name makes it clear.

3.  Argument **`xpath`**:

    -   **`xpath_expression`** - A more descriptive name that specifies the type of input expected.

    -   **`xml_path`** - A concise alternative that still hints at the purpose.

4.  Argument **`trim`**:

    -   **`remove_whitespace`** - A name that better reflects the action performed when **`trim`** is set to **`TRUE`**.

    -   **`clean_whitespace`** - Another option to indicate the removal of whitespace.

## Question 3

**Make a case for why `norm_r()`, `norm_d()` etc. would be better than [`rnorm()`](https://rdrr.io/r/stats/Normal.html), [`dnorm()`](https://rdrr.io/r/stats/Normal.html). Make a case for the opposite. How could you make the names even clearer?**

Function naming conventions in R often follow a pattern where functions are named with a verb followed by a noun, which makes the code more readable and easier to understand.

**Arguments for `norm_r()` and `norm_d()` (Custom Names):**

1.  **Verb-First Naming:** Putting the action (e.g., "normalize") before the distribution name (e.g., "norm") makes the function name more descriptive and aligns with R's naming conventions.

2.  **Consistency:** These names maintain consistency with other functions in R that follow a similar pattern (e.g., `read_csv()`, `plot_hist()`).

3.  **Clarity:** These names emphasize the action being performed (normalizing) over the distribution being used (normal distribution), which might be more intuitive to some users.

4.  **Readability:** These names read like a sentence: "Normalize a random sample from a normal distribution."

**Arguments for `rnorm()` and `dnorm()` (Current Names):**

1.  **Historical Convention:** Functions like `rnorm()` and `dnorm()` have been part of the base R distribution for a long time. Changing their names might lead to confusion for users who are already familiar with these functions.

2.  **Alignment with Mathematical Notation:** The `rnorm()` and `dnorm()` naming closely resembles mathematical notation, making it easier for users with a strong mathematical background to recognize and use these functions.

3.  **Package Consistency:** Maintaining the original names ensures consistency with other functions in the stats package (e.g., `pnorm()`, `qnorm()`), which also follow this convention.

**Even Clearer Names:** If we want to make the function names even clearer, we can consider longer, more descriptive names, such as:

-   `generate_random_samples_from_normal_distribution()` for `rnorm()`: This explicitly states what the function does.
-   `probability_density_function_of_normal_distribution()` for `dnorm()`: This specifies the exact purpose of the function.

While these longer names are more descriptive, they may become unwieldy in code. Striking a balance between clarity and brevity is important.
