---
title: "Chapter 9"
subtitle: "Workflow: getting help"
author: "Aditya Dahiya"
editor_options: 
  chunk_output_type: console
date: 2023-08-13
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/RfDS2solutions/Chapter9.html"
  description: "Solutions: R for Data Science (2e)"
  twitter: true
  facebook: true
  linkedin: true
  email: true
---

## 9.1 Google is your friend

We can copy any error; paste and search it on [google.com](www.google.com)

## 9.2 Making a `reprex`

Here, I try to create a reprex for a deliberate mistake I am making in `ggplot2` :\--

First, I create the faulty `ggplot2` code:\--

```{r}
#| label: ggplot2-faulty
#| error: false
#| warning: false
#| message: false
#| eval: false
library(tidyverse)
library(ggrepel)
library(gt)
data("gtcars")
gtcars |>
  slice_head(n = 3, 
             by = ctry_origin) |>
  ggplot(aes(x = hp,
             y = mpg_h,
             color = ctry_origin,
             label = model)) +
  geom_point() +
  geom_text_repel(force = 2) +
  theme_classic() +
  labs(x = "Horse Power",
       y = "Miles per Gallon (Highway)") +
  theme(legend.position = "bottom",
        legend.title = element_blank())
  
```

```{r}
#| label: ggplot2-reprex
#| error: false
#| warning: false
#| message: false
#| eval: false

# Copy the code above

# Then create reprex data set
gtcars |>
  slice_head(n = 3, 
             by = ctry_origin) |>
  dput()

```

Here's the reprex:

```{r}
#| eval: false

gtcars = 
structure(list(mfr = c("Ford", "Chevrolet", "Dodge", "Ferrari", 
"Ferrari", "Ferrari", "Acura", "Nissan", "Bentley", "Aston Martin", 
"Aston Martin", "BMW", "BMW", "BMW"), model = c("GT", "Corvette", 
"Viper", "458 Speciale", "458 Spider", "458 Italia", "NSX", "GT-R", 
"Continental GT", "DB11", "Rapide S", "6-Series", "i8", "M4"), 
    year = c(2017, 2016, 2017, 2015, 2015, 2014, 2017, 2016, 
    2016, 2017, 2016, 2016, 2016, 2016), trim = c("Base Coupe", 
    "Z06 Coupe", "GT Coupe", "Base Coupe", "Base", "Base Coupe", 
    "Base Coupe", "Premium Coupe", "V8 Coupe", "Base Coupe", 
    "Base Sedan", "640 I Coupe", "Mega World Coupe", "Base Coupe"
    ), bdy_style = c("coupe", "coupe", "coupe", "coupe", "convertible", 
    "coupe", "coupe", "coupe", "coupe", "coupe", "sedan", "coupe", 
    "coupe", "coupe"), hp = c(647, 650, 645, 597, 562, 562, 573, 
    545, 500, 608, 552, 315, 357, 425), hp_rpm = c(6250, 6400, 
    5000, 9000, 9000, 9000, 6500, 6400, 6000, 6500, 6650, 5800, 
    5800, 5500), trq = c(550, 650, 600, 398, 398, 398, 476, 436, 
    487, 516, 465, 330, 420, 406), trq_rpm = c(5900, 3600, 5000, 
    6000, 6000, 6000, 2000, 3200, 1700, 1500, 5500, 1400, 3700, 
    1850), mpg_c = c(11, 15, 12, 13, 13, 13, 21, 16, 15, 15, 
    14, 20, 28, 17), mpg_h = c(18, 22, 19, 17, 17, 17, 22, 22, 
    25, 21, 21, 30, 29, 24), drivetrain = c("rwd", "rwd", "rwd", 
    "rwd", "rwd", "rwd", "awd", "awd", "awd", "rwd", "rwd", "rwd", 
    "awd", "rwd"), trsmn = c("7a", "7m", "6m", "7a", "7a", "7a", 
    "9a", "6a", "8am", "8am", "8am", "8am", "6am", "6m"), ctry_origin = c("United States", 
    "United States", "United States", "Italy", "Italy", "Italy", 
    "Japan", "Japan", "United Kingdom", "United Kingdom", "United Kingdom", 
    "Germany", "Germany", "Germany"), msrp = c(447000, 88345, 
    95895, 291744, 263553, 233509, 156000, 101770, 198500, 211195, 
    205300, 77300, 140700, 65700)), class = c("spec_tbl_df", 
"tbl_df", "tbl", "data.frame"), row.names = c(NA, -14L), spec = structure(list(
    cols = list(mfr = structure(list(), class = c("collector_character", 
    "collector")), model = structure(list(), class = c("collector_character", 
    "collector")), year = structure(list(), class = c("collector_double", 
    "collector")), trim = structure(list(), class = c("collector_character", 
    "collector")), bdy_style = structure(list(), class = c("collector_character", 
    "collector")), hp = structure(list(), class = c("collector_double", 
    "collector")), hp_rpm = structure(list(), class = c("collector_double", 
    "collector")), trq = structure(list(), class = c("collector_double", 
    "collector")), trq_rpm = structure(list(), class = c("collector_double", 
    "collector")), mpg_c = structure(list(), class = c("collector_double", 
    "collector")), mpg_h = structure(list(), class = c("collector_double", 
    "collector")), drivetrain = structure(list(), class = c("collector_character", 
    "collector")), trsmn = structure(list(), class = c("collector_character", 
    "collector")), ctry_origin = structure(list(), class = c("collector_character", 
    "collector")), msrp = structure(list(), class = c("collector_double", 
    "collector"))), default = structure(list(), class = c("collector_guess", 
    "collector")), skip = 1), class = "col_spec"))

reprex::reprex()

```

````         
``` r
library(tidyverse)
library(ggrepel)

gtcars |>
  slice_head(n = 3, 
             by = ctry_origin) |>
  ggplot(aes(x = hp,
             y = mpg_h,
             color = ctry_origin,
             label = model)) +
  geom_point() +
  geom_text_repel(force = 2) +
  theme_classic() +
  labs(x = "Horse Power",
       y = "Miles per Gallon (Highway)") +
  theme(legend.position = "bottom",
        legend.title = element_blank())

```

<sup>Created on 2023-08-13 with [reprex v2.0.2](https://reprex.tidyverse.org)</sup>
````
